---
title: "lab3: Differetial Gene Expression Analysis"
author: "Your name or nickname"
date: "2025-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Lab workflow is based on the [DESeq2 tutorial](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html).

In this lab you get some hints and instructions. Once you solve the tasks, remove comment
symbols for lines of code, so that they are executed in the code chunks. Commenting you code
is important not only for future yourself, but also for other users to understand
what you have done. Please, provide meaningful comments for each step of the analysis.

```{r, message=FALSE}
# TODO: Load the necessary libraries for differential  gene expression analysis
# library(______) # tximport, check & write what this library does
# library(______) # tidyverse, check & write what this library does
# library(______) # DESeq2, check & write what this library does
# library(______) # AnnotationDbi, check & write what this library does
# library(______) # org.Hs.eg.db

# 💡 HELP: Check what these libraries do
# ?DESeq2
# ?AnnotationDbi
```

## Load Gene Level Quantification and Design File from Previous Lab

```{r}
# TODO: Load the saved data objects from the previous lab
# HINT: Use readRDS() function
# HINT: Load both txi object and design table

# txi <- ______("../______/______")
# design_file <- ______("../______/______")

# 💡 HELP: Learn about loading R objects
# ?readRDS
```

```{r}
# TODO: Construct DESeqDataSet from the txi object and sample information
# HINT: Use DESeqDataSetFromTximport() function
# HINT: You need: txi object, colData (sample info), and design formula
# HINT: Design formula should be ~ Condition

# ddsTxi <- ______(______,
#                  colData = ______,
#                  design = ~ ______) 

# ddsTxi

# 💡 HELP: Learn about DESeq2 objects
# ?DESeqDataSetFromTximport
```

```{r}
# TODO: Display experimental conditions 
# HINT: Use colData() function

# ______(ddsTxi)
```

**🧠 Think:** What information does colData contain? Why is this important for the analysis?

## Set Reference Condition for Comparison

```{r}
# TODO: Extract the control condition name to set as reference
# HINT: Use the provided code structure
# HINT: We want to extract "Control" from the Condition column

# reference <- as_tibble(colData(ddsTxi)) %>% 
#   dplyr::select(______) %>% 
#   filter(Condition == "______") %>% 
#   unique() %>% 
#   unlist() %>% 
#   as.character()
# reference
```

```{r}
# TODO: Set Control as the reference condition
# HINT: Use relevel() function
# HINT: The reference should be the variable you created above

# ddsTxi$Condition <- ______(ddsTxi$Condition, ref=______)
```

**🧠 Think:** Why do we need to set a reference condition? How does this affect fold change calculations?

## Run Differential Expression Analysis

```{r, message = FALSE}
# TODO: Run the DESeq2 analysis
# HINT: Use DESeq() function on your ddsTxi object

# dds <- ______(______)

# 💡 HELP: Learn about DESeq2 function
# ?DESeq
```

## Filter Low Count Genes

```{r}
# TODO: Extract and display the count matrix
# HINT: Use counts() function and head() to show first few rows

# head(______(______))

# According to DESeq2 tutorial, the count of 10 is a reasonable choice. We will
# keep only counts if there are at least 10 of them in at least certain number
# of samples (which will be equal to the number of samples in the smallest 
# group / condition)

# Define smallest group size (condition with smallest number of samples)
# smallestGroupSize <- ______  # HINT: How many samples we have for the smallest group?

# TODO: Create filtering criteria
# HINT: Use rowSums() with counts(dds) >= 10
# HINT: Compare to smallestGroupSize

# keep <- rowSums(counts(dds) >= ______) >= ______
# SUBSET dds object to keep entries that satisfy the condition above
# dds <- dds[______,]
```
## Extract and Summarize Results

```{r}
# TODO: Extract results with significance threshold of 0.05
# HINT: Use results() function with alpha parameter

# res05 <- results(______, alpha=______)
# summary(______)
```

**🧠 Think:** What do the numbers in the summary tell you about your experiment?

## Visualize Results with MA Plot

```{r, fig.width=5, fig.height=5}
# TODO: Create MA plot to visualize results
# HINT: Use plotMA() function
# ______(______)

# 💡 HELP: Learn about MA plots
# ?plotMA
```

**🧠 Think:** Inspect MA plot visually: what does it tell you?

## Create Results Directory

```{r}
# TODO: Create organized directory for saving results
# HINT: Use paste0() to combine strings
# HINT: Use dir.exists() and dir.create() to check and create directory

# project_name <- "______"
# result_folder <- ______("results_", ______)
# if(!______(result_folder)){
#   ______(result_folder)}

# 💡 HELP: Learn about directory operations
# ?dir.exists
# ?dir.create
```

## Save MA Plot

```{r}
# TODO: Save MA plot as PDF
# HINT: Use pdf() to open file, plotMA() to create plot, dev.off() to close
# HINT: Create meaningful filename using paste0()

# condition <- "EMT_vs_normal" # HINT: Describe your comparison (e.g., "Knockdown_vs_Control")
# ______(paste0(result_folder, "/", ______, "_MAplot.pdf")) 
# ______(res05)
# ______()

# 💡 HELP: Learn about saving plots
# ?pdf
# ?dev.off
# TODO: Check if the plot was saved in the specified location
```

## Principal Component Analysis

```{r, fig.width=5, fig.height=5}
# TODO: Apply variance stabilizing transformation
# HINT: Use vst() function with blind=FALSE
# vsd <- ______(dds, blind=______)

# TODO: Display sample information
# ______(vsd)
```

```{r, fig.width=5, fig.height=5}
# TODO: Create PCA plot
# HINT: Use plotPCA() with intgroup parameter
# HINT: Add a title using ggtitle()
# PCA_plot <- ______(vsd, intgroup=c("______")) + 
#   ggplot2::______("______") 
# PCA_plot

# 💡 HELP: Learn about variance stabilization and PCA
# ?vst
# ?plotPCA
```

**🧠 Think:** Do samples cluster by treatment condition? 

## Save PCA Plot 

```{r}
# TODO: Save PCA plot as PDF
# HINT: Follow same pattern as MA plot saving
# condtion <- "______"
# pdf(______(______, "/", ______, "_PCA___.pdf")) 
# PCA_plot
# dev.off()
# TODO: Check if the plot was saved in the specified location
```



## Add Gene Annotations

```{r}
# mapIds() maps Ensembl gene IDs to human-readable gene symbols

# TODO: Map Ensembl IDs to gene symbols
# HINT: Use mapIds() function from org.Hs.eg.db
# HINT: Parameters: database, keys, column, keytype, multiVals

# Add Gene Symbols
# res05$symbol <- ______(______,
#                        keys=______(______),
#                        column="SYMBOL",
#                        keytype="ENSEMBL",
#                        multiVals="first")
# res05

# 💡 HELP: Learn about gene annotation
# ?mapIds
# What is the meaning of multiVals argument and what is its default behaviour?
```

```{r}
# TODO: Add full gene descriptions
# HINT: Similar to above but use column="GENENAME"

# res05$gene_name <- ______(______,
#                           keys=______,
#                           column="______",
#                           keytype="______",
#                           multiVals="first")
# res05
```

```{r}
# TODO: Store Ensembl IDs in a column (at the moment Ensembl IDs are the rows)
# res05$ENSEMBL <- ______(res05)
```


```{r}
# TODO: Add Entrez Gene IDs for pathway analysis we do later
# HINT: Use column="ENTREZID"
# res05$ENTREZID <- mapIds(______,
#                           keys=______,
#                           column="______",
#                           keytype="______",
#                           multiVals="first")
```

```{r}
# Display annotated results & inspect additional annotation we added to our results
# res05
```

**🧠 Think:** Why do we need multiple types of gene identifiers? When would you use each type?

### Order and Export Results

```{r}
# order() sorts genes by statistical significance (lowest p-value first)

# TODO: Order results by statistical significance
# HINT: Use order() function on p-adjusted values

# resOrdered <- res05[______(res05$______),]
# 
# # Preview the results
# ______(resOrdered, 20)
```

```{r}
# as.data.frame() converts DESeq2 results to standard data frame for export

# TODO: Convert to data frame and save all results
# HINT: Use as.data.frame() and write_tsv()

# resOrderedDF <- ______(resOrdered)

# Save results as tsv file
# ______(resOrderedDF, 
#           file = paste0(______, "/", 
#                         ______, 
#                         "_all_genes_results.tsv"))

# TODO: Extract and save upregulated genes
# HINT: Use subset() with padj < 0.05 AND log2FoldChange > 0
# HINT: Use arrange() to sort by log2FoldChange (descending order)

# upreg_and_signif <- ______(
#   resOrderedDF,______ < ______ & ______ > ______) %>% 
#   ______(______(______))

# Preview table 
# ______(upreg_and_signif)

# Save results as tsv file
# ______(upreg_and_signif, 
#           file = paste0(______, "/", 
#                         ______, 
#                         "_upreg_genes_results.tsv"))

# TODO: Extract and save downregulated genes  
# HINT: Similar to above but log2FoldChange < 0
# HINT: Sort by log2FoldChange in ascending order

# downreg_and_signif <- ______(
#   resOrderedDF,______ < ______ & ______ < ______) %>% 
#   ______(______)

# Preview results 
# ______(downreg_and_signif)

# Save results as tsv file
# ______(______, file = paste0(______, "/", ______, "_downreg_genes_results.tsv"))
```

```{r}
# TODO: Record session information for reproducibility
# HINT: Use sessionInfo() function
```

**Answer these questions in your report:**

1. How many genes were significantly up- and down-regulated?

2. Do your MA plot and PCA plot suggest the experiment worked well? 

3. Look at your top differential expressed genes - do any make biological sense for your experimental setup? Report your observations.

4. Why do we have p-adjusted values in DESeq report?