---
title: "lab2: Load Data and Do Exploatory Data Analysis"
author: "Your name or nickname"
date: "2025-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lab you get some hints and instructions. Once you solve the tasks, remove comment
symbols for lines of code, so that they are executed in the code chunks. Commenting you code
is important not only for future yourself, but also for other users to understand
what you have done. Please, provide meaningful comments for each step of the analysis.

```{r}
# Load necessary libraries
# library(______) # tximport, check & write what this library does
# library(______) # tidyverse,check & write what this library does

# 💡 HELP: Check what these libraries do
# ?tximport
# ?tidyverse
```

## Project name 

Write short project description

## Data download

```{bash, eval = F}
# create a folder to store the data
# mkdir -p ../../data/{project_name}
```

```{bash, eval = F}
# data download 
#wget "https://www.dropbox.com/scl/fo/i387hjzocw227bjllf069/AB0Y-Rt3DxCAXMCJKJVGppM?rlkey=4feunl032pfpbr69w9yfj4al4&st=l2deogkv&dl=1" -O  #../../data/{project_name}/{project_data}.zip
```

**Unzip the downloaded files in the Terminal window of your RStudio.** 

```
unzip ../../data/{project_name}/{project_data}.zip -d ../../data/{project_name}/{project_data}
```

⚠️ Note: In the code chunks above, we used bash commands to create a folder where we download the data, which was specified by {bash}. Make sure eval = F is specified in these chunks (e.g., {bash, eval = F}) before rendering HTMLs, so that you do not do this repeatedly every time you perform a new rendering. You can also add eval = F to any chunk you don't want to execute when rendering the Rmd again.

## Create a design table from your experimental design info

```{r}
# TODO: Create a design table with sample information
# HINT: Use data.frame() function
# Your table should have two columns:
# 1. Sample_ID: the SRR numbers
# 2. Condition: depending on the expereimental design

# design_table <- ______(
#   Sample_ID = c("______", "______", "______", "______"),
#   Condition = c("______", "______", "______", "______")
# )
# design_table

# 💡 HELP: Learn about data frames
# ?data.frame
```

**🧠 Think:** Why is it important to have an experimental design table? What information does it capture?

## Load your data into R

```{r}
# TODO: Define the path to your data
# HINT: Look at the file structure from the download section

# Define base paths
# data_path <- "../../data/EMT"
# project <- "______"  # HINT: Look at the folder name with data

# paste0() concatenates strings to create file paths for data location
# data_path <- "../../data/EMT"
# project <- "TGFbeta_data"

# list.files() displays all files in the salmon quantification directory
# ________(paste0(data_path, "/", ______, "/", "salmon_quantification"))

# 💡 HELP: Understand file path construction
# ?paste0
# ?list.files
```

**🧠 Think:** What do you see when you list the files? These should be your sample IDs!

```{r}
# TODO: Create full file paths to each sample's quant.sf file
# HINT: Use file.path() function 
# HINT: Each sample has a folder named with its Sample_ID
# HINT: Inside each folder is a file called "quant.sf"

# file.path() constructs proper file paths to salmon quantification files for each sample

# quant_files <- ______(data_path, 
#                       ______, 
#                       paste0("salmon_quantification", "/", design_table$______), 
#                       "______")
# quant_files
```

```{r}
# TODO: Give meaningful names to your file paths
# HINT: Combine condition and sample ID for clarity
# HINT: Use paste0() to combine design_table$Condition and design_table$Sample_ID
# HINT: Use names() function to assign names to your quant_files

# ______(quant_files) <- paste0(design_table$______, "_", design_table$______)
# quant_files
```

```{r}
# TODO: Read the transcript-to-gene mapping file
# HINT: Use read_csv() function
# HINT: The file is called "tx2gene.ensembl_custom.v102.csv"

# tx2gene <- ______("______")
# tx2gene
```

**🧠 Think:** Why do we need to convert from transcripts to genes? What's the difference?

```{r}
# TODO: Import the salmon quantification data and summarize at gene level
# HINT: Use tximport() function
# HINT: You need three parameters:
#   - files: your quant_files vector
#   - type: "salmon" (since we used salmon for quantification)
#   - tx2gene: your mapping table

# tximport() imports salmon quantification data and summarizes at gene level using transcript-to-gene mapping
# txi <- ______(______, type = 'salmon', tx2gene=______)

# 💡 HELP: Learn about tximport
# ?tximport
```

```{r}
# TODO: Examine the structure of your imported data
# HINT: Use str() function

#______(txi)
```

```{r}
# TODO: Look at the first few rows of the count matrix
# HINT: Use head() function on txi$counts

# ______(txi$______)
```

```{r}
# TODO: Convert to tibble and select one sample
# HINT: Use as_tibble() and select() functions
# HINT: Try selecting the first TGFbeta3 sample

#______(txi$counts) %>% select(______)

# 💡 HELP: Learn about data conversion and selection
# ?as_tibble
# ?select
```

## Basic Data Visualization

Now let's create some simple plots to understand our data better!

### Library Size Visualization

**📖 Background:** Library size represents the total number of reads sequenced for each sample. Ideally, all samples should have similar library sizes, indicating consistent sequencing depth.

```{r}
# TODO: Calculate total read counts for each sample
# HINT: Use colSums() function on the count matrix
# HINT: The count matrix is stored in txi$counts

# library_sizes <- ______(txi$______)

# TODO: Create a data frame for plotting
# HINT: You need two columns:
# 1. Sample names (use names() function)
# 2. Library sizes (the values you just calculated)

# Create data frame for plotting
# lib_df <- data.frame(
#   Sample = ______(library_sizes),
#   Library_Size = ______
# )

# Look at your data frame
# lib_df

# 💡 HELP: Learn about column operations
# ?colSums
# ?names

# TODO: Create a bar plot showing library sizes
# HINT: Use ggplot() with geom_bar()
# HINT: Map Sample to x-axis and Library_Size to y-axis

# ggplot() creates bar plot of library sizes
# ______(lib_df, aes(x = ______, y = ______)) +
#   geom_bar(stat = "identity") +
#   labs(title = "______", x = "______", y = "______") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 💡 HELP: Learn about ggplot2
# ?ggplot
# ?geom_bar
# ?labs
# ?theme
```

### Count Distribution

```{r}
# TODO: Transform count data to log scale
# HINT: Use log2() function
# HINT: Add 1 to avoid log(0) which is undefined
# HINT: Apply this to txi$counts

# log_counts <- log2(txi$______ + _)

# TODO: Convert to long format for ggplot
# HINT: Start with log_counts, convert to as_tibble
# HINT: Use pivot_longer() to reshape the data
# HINT: everything() selects all columns
# HINT: Set names_to = "Sample" and values_to = "Log_Counts"

# count_df <- log_counts %>%
#   as______() %>%
#   pivot_longer(______(), names_to = "______", values_to = "______")

# Look at the first few rows
# ______(count_df)

# TODO: Create histogram showing count distributions
# HINT: Use ggplot() with geom_histogram()
# HINT: Map Log_Counts to x-axis
# HINT: Use facet_wrap() to create separate panels for each sample
# HINT: Set bins = 20 for histogram

# ggplot() creates histogram of log-transformed counts
# ______(count_df, aes(x = ______)) +
#   geom_histogram(bins = __) +
#   facet_wrap(~Sample) +
#   labs(title = "______", x = "______", y = "______")

# 💡 HELP: Learn about histograms and faceting
# ?geom_histogram
# ?facet_wrap
```

**🧠 Think:**
- Do all samples show similar distribution shapes?
- What does the peak around 0 represent?
- Why do we see a long tail to the right?

## Save Data for the Next Analysis Steps

**📖 Background:** RDS files preserve R objects exactly as they are, making them perfect for saving complex data structures for later analysis.

```{r}
# TODO: Save the tximport object for future use
# HINT: Use saveRDS() function
# HINT: Save txi object as "txi_EMT_data.rds"

# saveRDS() saves the tximport object as RDS file for use in downstream analysis
# ______(txi, file = "______.rds")

# Also save design table for reference
# saveRDS(______, file = "design_table.rds")
```

**Answer these questions in your report:**

1. **Data Structure:** What does the `txi` object contain? What are the different components (hint: use `names(txi)`)?

2. **Sample Naming:** Why did we create meaningful names for our samples? How does this help in downstream analysis?

3. **File Organization:** How is the salmon output organized? What would happen if the file paths were incorrect?

## Troubleshooting Tips

**Common Issues:**

1. **File not found errors:** 
   - Check your file paths carefully
   - Make sure you've unzipped the data
   - Use `list.files()` to verify file locations

2. **Object not found errors:**
   - Check that you've created all required objects
   - Make sure you've run all previous code chunks

3. **Function not found errors:**
   - Check that you've loaded all required libraries
   - Use `?function_name` to verify function exists

4. **Column name errors:**
   - Check your design table column names
   - Make sure Sample_ID and Condition are spelled correctly
   
**🎉 Congratulations!** You've successfully imported count data and saved it for downstream analysis!
